\documentclass[sigconf,nonacm,screen]{acmart}
\settopmatter{printfolios=true}

% --- Packages beyond acmart's own (acmart loads amsmath/amsthm, natbib,
% --- hyperref, graphicx, xcolor, booktabs, microtype) ---
\usepackage{balance}
\usepackage{pgfplots}
\pgfplotsset{compat=1.16}
\usepackage{tikz}
\usetikzlibrary{positioning,fit,arrows.meta,calc}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{listings}
\usepackage{cleveref}

% ============================================================
% Measured numbers. RULE: a macro may cite only a committed
% bench/results/*.json. The generated file is produced by
% `make paper-data` and verified byte-identical by
% `make paper-check`; hand edits to it do not survive.
% Hardware: Apple M5 Max (40-core GPU, 128 GB unified memory)
% unless a macro's comment says otherwise.
% ============================================================
\input{generated/results-macros}

% --- Shared figure style: Okabe-Ito palette, one pgfplots style ---
\definecolor{kippblue}{HTML}{0072B2}
\definecolor{kipporange}{HTML}{E69F00}
\definecolor{kippgreen}{HTML}{009E73}
\definecolor{kippverm}{HTML}{D55E00}
\definecolor{kipppurple}{HTML}{CC79A7}
\pgfplotscreateplotcyclelist{kippcycle}{%
  {kippblue, mark=*},
  {kipporange, mark=square*},
  {kippgreen, mark=triangle*},
  {kippverm, mark=diamond*},
  {kipppurple, mark=pentagon*}}
\pgfplotsset{kippplot/.style={
  width=\columnwidth, height=4.6cm, tick align=inside,
  ymajorgrids, grid style={gray!25},
  legend style={draw=none, font=\footnotesize},
  cycle list name=kippcycle,
  every axis plot/.append style={line width=0.9pt, mark size=1.6pt}}}

\lstset{basicstyle=\ttfamily\footnotesize,breaklines=true,frame=single,
        columns=fullflexible,keepspaces=true}
\newcommand{\code}[1]{\texttt{\small #1}}

\begin{document}

\title{Kipp: Oracle-Gated, Placement-Invariant Paged Attention for
Verifiable LLM Inference}
\author{David Khachatryan Johansson}
\affiliation{%
  \institution{Cortify}
  \city{Stockholm}
  \country{Sweden}}
\email{david@cortify.io}

\begin{abstract}
Paged key--value (KV) caches are standard in LLM serving, and every system
that relocates cache blocks at runtime---for prefix sharing, migration,
spilling, or speculative rollback---relies on an assumption none of them
tests: the physical placement of KV blocks must not change the model's
output. We turn that assumption into a machine-checked contract. Placement
invariance is stated as a proposition whose hypotheses map to real bug
classes and enforced by a \emph{scramble gate} that reverses a sequence's
block table and requires bitwise-identical logits. A mutation study shows
why tolerance-based reference testing is not enough: of four seeded,
realistic KV-addressing faults, a block-rollover bug measures NMSE of
exactly $0$ against the reference---invisible to any tolerance, however
tight---and only the scramble gate catches it, while two read-side faults
never flip the arg\,max and fall only to a tight NMSE bound, and a fourth
is gross corruption any check catches. The two gate families are
complementary, and the study also repaired a blind spot in
our own gate. We embody the discipline in Kipp, a hand-written C11 engine
for the Qwen3 dense family whose byte-stable scalar oracle gates two GPU
backends and every feature: 8- and 4-bit quantization, simdgroup-matrix
prefill, continuous batching, adaptive speculation, and pooled KV sharing.
Kipp wins bandwidth-bound decode against llama.cpp on identical hardware,
prefix reuse cuts time to first token by $\PrefixSpeedup\times$, and every
benchmark number in this paper is bound to a committed, re-runnable
artifact.
\end{abstract}

\maketitle

\section{Introduction}
\label{sec:intro}
Serving a large language model (LLM) is dominated by the cost of the
key-value (KV) cache. \citet{kwon2023pagedattention} observed that reserving
a contiguous KV region per request wastes $60$--$80\%$ of KV memory to
internal and external fragmentation, and introduced \emph{PagedAttention}:
store KV in fixed-size blocks and give each sequence a \emph{block table}
that maps logical positions to arbitrary physical blocks, exactly as an
operating system pages virtual memory. Paging is now standard practice
\citep{kwon2023pagedattention,zheng2024sglang,miao2024servingsurvey}, and it
underpins prefix sharing, live block migration, and speculative-decode
rollback.

Paging rests on an unstated assumption: where a block physically lives must
not change what the model computes. Systems that relocate KV at runtime take
this for granted, whether they migrate running requests between instances
\citep{sun2024llumnix}, spill cache entries between GPU and host memory
\citep{lee2024infinigen}, ship KV between prefill and decode clusters
\citep{qin2025mooncake}, persist it across conversation turns
\citep{gao2024cachedattention}, or share it across calls
\citep{zheng2024sglang}. Of the systems \Cref{tab:survey} surveys, ten
relocate KV through an indirection and none tests placement invariance
(the eleventh, vAttention, sidesteps the indirection by construction). However, the indirection they all depend on, a
per-access \code{block\_table[pos>>5]*32 + (pos\&31)} lookup replicated
independently in every backend's write and read kernels, is exactly where
off-by-one and stride bugs hide. Such bugs are silent. They corrupt a few
positions without crashing, and a tolerance-based comparison to a reference
forward pass, which is the standard test, can mask them as numerical noise
(\Cref{sec:silent}).

Kipp takes the opposite stance: correctness properties that serving systems
rely on should be adversarially tested, continuously, on every backend. This
paper makes the following contributions.

\begin{enumerate}\itemsep2pt
  \item We state placement invariance as a proposition with explicit
  hypotheses, each of which corresponds to a real bug class, and realize it
  as a \emph{scramble gate}: reverse a sequence's block table and require the
  output logits to be bitwise identical to the identity mapping
  (\Cref{sec:vpa}).
  \item We evaluate paging-test adequacy by mutation. Seeding \FaultCount{}
  realistic, asymmetric KV-addressing bugs shows that the detection pattern
  splits exactly along the proposition's hypotheses: placement-commuting
  faults are caught only by reference testing, two of them without ever
  flipping the arg\,max, and the placement-coupled rollover fault, which
  measures NMSE of exactly $0$ against the reference, is caught only by the
  scramble gate. The study also exposed a blind spot in our own gate and
  drove its fix (\Cref{sec:eval-faults}).
  \item We hold a scalar CPU forward pass byte-identical across the engine's
  evolution and use it as a hard gate: Metal and CUDA must match its
  arg\,max exactly and stay within $10^{-4}$ NMSE (\Cref{sec:oracle}).
  \item We build and evaluate the engine itself (\Cref{tab:loc}): a C11 core
  running the Qwen3 dense family with 8- and 4-bit quantization,
  simdgroup-matrix prefill, continuous batching, an OpenAI-compatible
  server, and prompt-lookup speculation with adaptive gating, every feature
  admitted through the same gates, and evaluated end to end, from
  microbenchmarks and quantization quality to serving under load and a
  same-hardware llama.cpp comparison (\Cref{sec:impl,sec:eval}).
  \item We bind every benchmark number in this paper to a committed result
  file produced by the repository's harness, and every correctness claim
  to a gate command listed in the artifact appendix; the two out-of-band
  measurements are marked as such where they appear.
\end{enumerate}

We are explicit about what is not new. Attention is invariant to a
permutation of stored KV under a matching permutation of the causal mask;
this is folklore and has been treated formally in recent KV-quantization
work. Zero-copy speculative rollback is standard paged-cache behavior
\citep{leviathan2023speculative}. Our contribution is the engineering
discipline of testing the assumption, the measurement of what existing test
regimes miss, and the artifact that results.

\section{Background}
\label{sec:bg}

\subsection{Transformer inference and the KV cache}
Autoregressive inference has two phases with opposite bottlenecks.
\emph{Prefill} evaluates the whole prompt at once and is compute-bound, since
every weight is reused across all prompt tokens. \emph{Decode} produces one
token per step and is bandwidth-bound: each step streams every weight from
memory to produce a single token, so throughput tracks bytes per weight.
This is why weight quantization accelerates decode but not prefill
(\Cref{sec:eval-throughput}). Each generated token appends a key and value
vector per layer to the KV cache, which grows linearly with context and
persists across steps.

Kipp targets a fixed registry of pinned Qwen3 dense checkpoints
(0.6B--32B, base and instruct) \citep{yang2025qwen3}. The family shares
head dimension $128$, eight KV heads \citep{ainslie2023gqa}, a vocabulary
of $151{,}936$, RMSNorm $\epsilon=10^{-6}$, NEOX rotary embeddings
\citep{su2024rope}, grouped-query attention, and SwiGLU MLPs; only depth,
width, query-head count, and rope~$\theta$ vary. Fixing the family, rather than accepting
arbitrary GGUF files, is what makes exact per-checkpoint validation
tractable.

\subsection{Paged KV caches and the OS analogy}
Kipp stores a sequence's KV in physical blocks of $B{=}32$ positions. A
per-sequence \emph{block table} $\pi$ maps logical block $b$ to a physical
block $\pi(b)$; position $p$ resolves to physical slot
$\pi(\lfloor p/B\rfloor)\cdot B + (p \bmod B)$. The analogy to OS paging is
exact: blocks are pages, the block table is a page table, and logical
positions are virtual addresses. The identity table $\pi(b){=}b$ reproduces a
contiguous cache byte for byte. Prefix sharing, block migration, and rollback
are all realized by editing $\pi$, never by moving KV bytes. The analogy also
suggests the missing test. An OS would not ship an MMU whose translation path
had only ever been exercised with identity mappings, yet that is how paged
LLM engines are tested today.

\subsection{Speculation on bandwidth-bound hardware}
Because decode streams every weight per step, a $k{+}1$-token speculative
verify forward is nearly free on compute-bound hardware but costs close to
$k{+}1$ decode steps when weight streaming dominates
\citep{leviathan2023speculative,saxena2023promptlookup}. Speculation is
therefore a net loss at low acceptance, which motivates the adaptive gate of
\Cref{sec:impl-spec}. The same arithmetic makes batched decode scale:
one weight read serves every concurrently decoding sequence
\citep{yu2022orca}, which \Cref{sec:eval-batching} measures.

\section{The Silent-Corruption Problem}
\label{sec:silent}
The KV indirection is two instructions, but several distinct bug classes
live in it:

\begin{itemize}\itemsep2pt
  \item \emph{Block-index errors}: an off-by-one in the logical-block
  computation reads a neighboring block's KV.
  \item \emph{Slot errors}: an off-by-one within the block shifts every
  source position by one slot.
  \item \emph{Rollover errors}: the first position after a block boundary is
  addressed relative to the previous block, one past its end.
  \item \emph{Binding errors}: key and value stores, or their strides, are
  swapped in one kernel's argument list.
\end{itemize}

Three properties make these bugs dangerous. First, they are silent. The
wrong address is still a valid slot, so no bounds are violated, nothing
crashes, and generation continues fluently; attention's softmax dilutes a few
corrupted source positions into a plausible output distribution. Second,
they are masked by tolerance. The standard test compares final logits to a
reference within a numerical tolerance chosen to absorb legitimate
reduction-order and quantization noise, and a bug that corrupts a few
positions can land within or near that band, indistinguishable from noise.
Third, some are invisible under the identity mapping. A rollover bug that
writes position $32b$ ``one past the end of block $b{-}1$'' lands on exactly
the right byte when blocks are laid out contiguously. Under the identity
table every happy-path test passes bitwise, and the bug only corrupts once a
scheduler actually relocates a block. \Cref{sec:eval-faults} demonstrates
each of these behaviors with seeded faults.

The exposure grows with exactly the features that make paging valuable:
cross-request prefix sharing backs the same physical block with many
sequences, defragmentation compacts live blocks with zero recompute, and
speculative rollback drops rejected blocks by unlinking. Each new feature
adds code that manipulates $\pi$. None of it is exercised by tests that only
ever run the identity mapping. \Cref{tab:survey} makes the claim auditable:
across ten widely used systems that relocate or reuse KV at runtime, we
found no test of placement invariance as a property.

\begin{table}[t]\centering\small
\caption{Systems that relocate or reuse KV at runtime, and whether their
test suites check placement invariance as a property. Based on inspection of
each system's public repository and test suite as of July 2026; we welcome
corrections.}
\label{tab:survey}
\begin{tabular}{@{}lp{3.55cm}c@{}}
\toprule
System & KV relocation / reuse & Tests inv.? \\
\midrule
vLLM \citep{kwon2023pagedattention} & block-table paging, prefix sharing & no \\
SGLang \citep{zheng2024sglang} & radix-tree prefix sharing & no \\
TensorRT-LLM \citep{nvidia2023tensorrtllm} & paged KV, block reuse & no \\
TGI \citep{huggingface2023tgi} & paged KV & no \\
llama.cpp \citep{llamacpp2023gguf} & unified KV cells, defrag moves & no \\
Mooncake \citep{qin2025mooncake} & prefill$\to$decode KV transfer & no \\
InfiniGen \citep{lee2024infinigen} & GPU--host KV spilling & no \\
Llumnix \citep{sun2024llumnix} & live request + cache migration & no \\
CachedAttention \citep{gao2024cachedattention} & cross-turn KV persistence & no \\
vAttention \citep{prabhu2025vattention} & contiguous virtual KV, demand paging & n/a$^{a}$ \\
\midrule
Kipp (this work) & block-table paging, pooled prefix sharing &
  \textbf{yes}$^{b}$ \\
\bottomrule
\multicolumn{3}{@{}l}{\scriptsize $^a$ Avoids table indirection by
construction (contiguous virtual addresses).} \\
\multicolumn{3}{@{}l}{\scriptsize $^b$ Bitwise scramble gate on CPU and
Metal (\Cref{sec:vpa-gate}).}
\end{tabular}
\end{table}

\section{Placement-Invariant Paged Attention}
\label{sec:vpa}

\subsection{The invariance property}
\label{sec:vpa-prop}

\begin{proposition}[Physical-placement invariance]
\label{prop:invariance}
Fix a sequence of logical positions $0,\dots,L-1$ with token embeddings,
rotary phases, and causal mask defined on logical positions. Let the KV cache
be stored in physical blocks of $B$ slots addressed through a block table,
and let $\pi,\pi'$ be any two injective maps from logical blocks to physical
blocks. Suppose:
\begin{itemize}\itemsep1pt
  \item[\textbf{H1}] the same map is applied to every KV \emph{write} and
  every KV \emph{read} (the address is
  $\pi(\lfloor p/B\rfloor)\cdot B + (p\bmod B)$, and nothing else in the
  kernel depends on the physical address);
  \item[\textbf{H2}] rotary embeddings and the causal mask are computed from
  logical positions only; and
  \item[\textbf{H3}] the attention reduction enumerates source positions
  in a fixed logical order, with a fixed summation order and the same
  floating-point operations, independent of $\pi$.
\end{itemize}
Then for every query position the attention output, and hence the final
logits, computed under $\pi$ and $\pi'$ are bitwise identical.
\end{proposition}

\begin{proof}[Proof sketch]
Under H1 the bytes written for logical position $p$ and the bytes read for
$p$ coincide regardless of $\pi$: read$\,\circ\,$write is the identity on
logical positions, so by H2 the sequence of (key, value, phase, mask) tuples
entering attention at any query position is equal as an indexed sequence,
not merely as a set. By H3 the reduction consumes that sequence in the same
order with the same operations, and floating-point evaluation is a
deterministic function of operand bits and operation order, so every
intermediate value is bit-equal. Induction over layers extends equality to
the final logits.
\end{proof}

Each hypothesis is load-bearing, and each corresponds to a real bug class.
Violating H1 on one side of the indirection is exactly the fault family of
\Cref{sec:silent}; a mapping error applied consistently to both sides is a
harmless relocation, which is why our mutation study must (and does) seed
one-sided faults. H2 fails if any kernel derives a rotary phase or mask
bound from a physical address. H3 is the batch-invariance axis of
\citet{he2025defeating}. The proposition does not require determinism across
batch sizes, only that placement not perturb the reduction; the two axes are
orthogonal, and a kernel can satisfy either without the other.

\subsection{Scope, stated precisely}
Invariance holds for the physical placement of retained KV, that is, for
which physical block holds a given logical position. It does not hold for
reordering the logical positions themselves, since rotary embeddings and the
causal mask are position-dependent. Nor is the triangular prefill mask
invariant to permuting columns of the score matrix. We test and claim only
physical-placement invariance. This distinction is what keeps the contract
honest.

\subsection{The scramble gate}
\label{sec:vpa-gate}
We realize the contract as an adversarial test (\Cref{lst:gate},
\Cref{fig:scramble}). The gate builds a sequence spanning several blocks,
evaluates it under the identity table, then reverses the block table (a
non-identity permutation that relocates every logical block) and re-evaluates
the same tokens. The final logits must be \code{memcmp}-equal. In the
terminology of software testing this is a metamorphic test
\citep{chen2018metamorphic}: the block-table permutation is a metamorphic
relation whose expected effect on the output is exactly zero. Because a
scrambled table exercises the addressing arithmetic with a mapping that no
happy-path test would produce, the gate tests the contrapositive of
\Cref{prop:invariance}. Any bitwise difference falsifies one of H1--H3 and
localizes a bug that a reference-NMSE comparison could report as $10^{-3}$
``noise.'' The gate runs on both the CPU oracle (\code{--paged-cpu}) and the
Metal backend (\code{--paged-metal}); it is cheap (\Cref{sec:oracle}) and
backend-agnostic.

\begin{lstlisting}[caption={The placement-invariance gate (paraphrased).},
                   label={lst:gate}]
eval(identity_session, tokens -> logits_A)
reverse(scrambled_session.block_table)   // non-identity permutation
eval(scrambled_session, tokens -> logits_B)
assert memcmp(logits_A, logits_B) == 0   // bitwise, not tolerance
\end{lstlisting}

\begin{figure}[t]\centering
\begin{tikzpicture}[
  blk/.style={draw, minimum width=0.72cm, minimum height=0.5cm,
              font=\scriptsize, inner sep=1pt},
  lbl/.style={font=\scriptsize}, >=Stealth]
  % identity physical row (above), logical row (middle), reversed (below):
  % arrows fan out from the logical row without crossing any node row.
  \node[lbl] at (-1.15,1.1) {$\pi=\mathrm{id}$};
  \foreach \i/\c in {0/kippblue!18,1/kippgreen!20,2/kipporange!28,3/kippverm!18} {
    \node[blk, fill=\c] (P\i) at (\i*0.85, 1.1) {$\i$};
  }
  \node[lbl] at (-1.15,0) {logical};
  \foreach \i/\c in {0/kippblue!18,1/kippgreen!20,2/kipporange!28,3/kippverm!18} {
    \node[blk, fill=\c] (L\i) at (\i*0.85, 0) {$b_\i$};
    \draw[->, gray] (L\i.north) -- (P\i.south);
  }
  \node[lbl] at (-1.15,-1.25) {$\pi=\mathrm{rev}$};
  \foreach \i/\j/\c in {0/3/kippblue!18,1/2/kippgreen!20,2/1/kipporange!28,3/0/kippverm!18} {
    \node[blk, fill=\c] (R\i) at (\j*0.85, -1.25) {$\j$};
    \draw[->, gray] (L\i.south) to[out=-90, in=90, looseness=0.7]
      (R\i.north);
  }
  \node[lbl, align=center] at (1.3,-2.05)
    {same tokens, same logical order, relocated bytes\\
     $\Rightarrow$ \code{memcmp}(logits) $= 0$};
\end{tikzpicture}
\caption{The scramble gate. The same four logical blocks are evaluated under
the identity table and under a reversed table that relocates every block;
color tracks the logical block. A correct engine produces bitwise-identical
logits, and any one-sided addressing bug produces a bitwise difference.}
\Description{Diagram of four logical KV blocks mapped once by the identity
block table and once by a reversed block table.}
\label{fig:scramble}
\end{figure}

\subsection{The gate lattice}
\label{sec:vpa-lattice}
The scramble gate is one node in a lattice of gates that every change must
pass (\Cref{fig:lattice}). Pinned FP32 reference vectors anchor the CPU
oracle at exact arg\,max and NMSE $\le 5\times10^{-5}$ (the oracle's own
BF16 KV-storage contract rules out bitwise equality with an FP32
reference; what \emph{is} bitwise is the oracle's stability across engine
versions, \Cref{sec:oracle}). The paged gates anchor both backends' KV
addressing bitwise. The
cross-backend gates anchor Metal and CUDA to the oracle at exact arg\,max and
NMSE $\le 10^{-4}$, and the speculation gate requires speculative output to
be token-identical to plain greedy decoding. A feature is not merged until
the whole lattice is green on real hardware. Re-running the lattice on every
engine change is tenable because the gates are fast (\Cref{tab:gatecost}).

\begin{figure}[t]\centering
\begin{tikzpicture}[
  gate/.style={draw, rounded corners=1pt, align=center,
               font=\scriptsize, inner sep=3pt},
  crit/.style={font=\tiny\itshape, midway, fill=white, inner sep=1pt},
  key/.style={draw, minimum width=0.28cm, minimum height=0.28cm, inner sep=0pt},
  >=Stealth]
  % Left-to-right: anchoring vectors -> oracle -> backends -> derived gates.
  % Fill encodes the criterion class of the node's incoming edge.
  \node[gate, fill=kippblue!14] (vec) at (0,0) {pinned\\vectors};
  \node[gate, fill=kippblue!14] (cpu) at (1.75,0) {CPU\\oracle};
  \node[gate, fill=kippblue!14] (pagedc) at (3.9,1.0) {paged CPU\\(scramble)};
  \node[gate, fill=kipporange!22] (metal) at (3.9,-0.1) {Metal};
  \node[gate, fill=kipporange!22] (cuda) at (3.9,-1.2) {CUDA};
  \node[gate, fill=kippblue!14] (pagedm) at (6.3,-0.1) {paged Metal\\(scramble)};
  \node[gate, fill=kippgreen!16] (spec) at (6.3,1.0) {speculative\\decode};
  \draw[->] (vec) -- node[crit] {arg\,max, $5{\cdot}10^{-5}$} (cpu);
  \draw[->] (cpu) -- node[crit, pos=0.55] {bitwise} (pagedc);
  \draw[->] (cpu) -- node[crit] {arg\,max, $10^{-4}$} (metal);
  \draw[->] (cpu) -- node[crit, pos=0.6] {arg\,max, $10^{-4}$} (cuda);
  \draw[->] (metal) -- node[crit] {bitwise} (pagedm);
  \draw[->] (pagedc) -- node[crit, yshift=8pt, fill=none] {token-identical} (spec);
  % criterion-class legend
  \node[key, fill=kippblue!14]  (k1) at (0.1,-1.35) {};
  \node[font=\tiny, right=1pt of k1] (k1t) {bitwise};
  \node[key, fill=kipporange!22, right=8pt of k1t] (k2) {};
  \node[font=\tiny, right=1pt of k2] (k2t) {arg\,max + NMSE};
  \node[key, fill=kippgreen!16, below=4pt of k1] (k3) {};
  \node[font=\tiny, right=1pt of k3] {token-identical};
\end{tikzpicture}
\caption{The gate lattice, left to right from anchor to derived gates. Every
edge is a machine-checked comparison; node fill encodes the criterion class
of its incoming edge. A change merges only when the lattice is green on real
hardware. CUDA's node covers its actual scope: contiguous KV, BF16 weights,
final-row logits.}
\Description{Directed graph of correctness gates from pinned vectors through
the CPU oracle to backend and derived gates.}
\label{fig:lattice}
\end{figure}

\section{Oracle-Gated Determinism}
\label{sec:oracle}
Kipp keeps a scalar CPU forward pass whose BF16 output is byte-identical
across the engine's entire evolution: the same logits, bit for bit, as the
first release. It is deliberately un-optimized. Its only job is to be a
stable oracle. Every other backend is gated against it, with identical
arg\,max at prefill and every checked decode position, and full-logit NMSE
at most $10^{-4}$. Quantized weight schemes are gated the same way against
the BF16 oracle, with exact arg\,max and a scheme-appropriate NMSE bound.

The kernel-verification literature commonly assumes that optimized GPU
kernels are essentially never bitwise-equal to a reference, so one must
``model floats as reals'' and check within tolerance. Floating-point
summation is indeed non-associative, and reduction-order differences are a
documented source of irreproducibility in both HPC and deep learning
\citep{shanmugavelu2024fpna}. We invert the assumption instead of accepting
it: a scalar reference held byte-stable serves as a hard gate, and a real
multi-backend engine holds its GPU paths to exact arg\,max and near
machine-precision NMSE against it. The discipline is differential testing
\citep{mckeeman1998differential}, applied across backends as CRADLE applies
it across deep-learning frameworks \citep{pham2019cradle}, with one twist:
the reference is not another implementation but a frozen, readable version
of this one, which makes every deviation attributable.

Verification is cheap enough to run continuously. \Cref{tab:gatecost} lists
the wall-clock cost of each gate on the evaluation machine. The entire
lattice costs about \GateCostTotalMinutes{}~minutes, dominated by the scalar
oracle's paged and pooled gates; every GPU-side gate runs in seconds to a
few minutes.

\begin{table}[t]\centering\small
\caption{Gate costs (Qwen3-4B, bf16, single run, idle machine). ``bitwise''
= \code{memcmp}-equal full-vocabulary logits. Values from
\code{bench/results/gate-costs.json}.}
\label{tab:gatecost}
\begin{tabular}{llr}
\toprule
Gate & Criterion & Cost (s) \\
\midrule
\code{--model} (CPU vs.\ pinned)        & NMSE $\le 5{\cdot}10^{-5}$, arg\,max & \GateCostModel \\
\code{--multilogit}                     & bitwise                              & \GateCostMultilogit \\
\code{--phase2-model} (KV cache)        & NMSE $\le 10^{-6}$, arg\,max         & \GateCostPhaseTwoModel \\
\code{--paged-cpu} (scramble)           & bitwise                              & \GateCostPagedCpu \\
\code{--pooled-cpu} (sharing)           & bitwise                              & \GateCostPooledCpu \\
\code{--metal-operators}                & exact per-kernel                     & \GateCostMetalOperators \\
\code{--phase3-metal} (Metal vs.\ CPU)  & NMSE $\le 10^{-4}$, arg\,max         & \GateCostPhaseThreeMetal \\
\code{--paged-metal} (scramble)         & bitwise                              & \GateCostPagedMetal \\
\code{--multilogit-metal}               & NMSE $\le 10^{-4}$, arg\,max         & \GateCostMultilogitMetal \\
\code{--pooled-metal} (sharing)         & bitwise + oracle NMSE                & \GateCostPooledMetal \\
\bottomrule
\end{tabular}
\end{table}

\section{Engine Design and Implementation}
\label{sec:impl}

\subsection{Architecture}
Kipp is small enough to read (\Cref{tab:loc}, \Cref{fig:arch}); Objective-C
is confined to the Metal bridge and CUDA to its kernels. A single core owns
strict GGUF validation against a compiled-in checkpoint registry, the native
Qwen tokenizer, sampling, session lifecycle, and the scalar reference
forward pass. Backends implement one synchronous \code{eval} contract
(append tokens to a session, produce logits for the last $n$ positions) and
own their weight representation, activation scratch, and KV storage. The
loader \code{mmap}s weights from a GGUF-style container and never copies
them fully into RAM, so a 4B model serves in about \PeakRSS~MB of resident
memory. No backend falls back to another, and a model outside the registry
is rejected rather than accommodated.

\begin{table}[t]\centering\small
\caption{Implementation size by component (\code{wc -l}, July 2026).}
\label{tab:loc}
\begin{tabular}{@{}llr@{}}
\toprule
Component & Language & Lines \\
\midrule
Core: loader, tokenizer, oracle, sessions, & & \\
\quad block pool, chat, speculation & C11 & 5{,}370 \\ % wc -l, 2026-07-21
Metal backend (bridge + kernels) & ObjC + MSL & 3{,}128 \\ % wc -l, 2026-07-21
CUDA backend & CUDA C++ & 1{,}587 \\ % wc -l, 2026-07-21
Server & C11 & 3{,}017 \\ % wc -l, 2026-07-21
CLI & C11 & 694 \\ % wc -l, 2026-07-21
Tests and gates & C11 + Python & 3{,}602 \\ % wc -l, 2026-07-21
Benchmark harness & Python & 1{,}346 \\ % wc -l, 2026-07-21
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[t]\centering
\begin{tikzpicture}[
  box/.style={draw, rounded corners=1pt, align=center, font=\scriptsize,
              inner sep=3pt},
  core/.style={box, fill=kippblue!8}, be/.style={box, fill=kipporange!12},
  gatebox/.style={box, fill=kippgreen!10}, >=Stealth]
  \node[core, minimum width=6.6cm] (server)
    {OpenAI server (3.0\,kLoC): completions, chat, SSE, metrics\\
     FIFO scheduler, chunked prefill, continuous batching};
  \node[core, minimum width=6.6cm, below=0.28cm of server] (corec)
    {C11 core (5.4\,kLoC): registry-validated GGUF loader, tokenizer,\\
     sampler, sessions, paged-KV block tables, block pool};
  \node[be, below left=0.5cm and -2.35cm of corec] (cpu)
    {CPU oracle\\ \emph{scalar, byte-stable}};
  \node[be, right=0.28cm of cpu] (metal)
    {Metal (3.1\,kLoC)\\ \emph{paged, quantized, MMA}};
  \node[be, right=0.28cm of metal] (cuda)
    {CUDA (1.6\,kLoC)\\ \emph{contiguous, BF16}};
  \node[gatebox, minimum width=6.6cm, below=1.5cm of corec] (gates)
    {gate lattice (\Cref{fig:lattice}): bitwise + arg\,max + NMSE,\\
     run on real hardware before any feature is ``done''};
  \draw[->] (server) -- (corec);
  \draw[->] (corec.south -| cpu) -- (cpu.north);
  \draw[->] (corec.south -| metal) -- (metal.north);
  \draw[->] (corec.south -| cuda) -- (cuda.north);
  \draw[->] (cpu.south) -- (gates.north -| cpu);
  \draw[->] (metal.south) -- (gates.north -| metal);
  \draw[->] (cuda.south) -- (gates.north -| cuda);
\end{tikzpicture}
\caption{Kipp's architecture. One core drives three backends through a
single \code{eval} contract, and every backend answers to the gate lattice.}
\Description{Layer diagram: server above a C11 core above three backends,
all feeding a gate-lattice box.}
\label{fig:arch}
\end{figure}

\subsection{Paged KV manager and block pool}
\label{sec:impl-paged}
Both the CPU oracle and the Metal backend address KV through the per-sequence
block table of \Cref{sec:vpa}. The physical store is a whole number of
32-position blocks, and the block-table lookup is a two-instruction address
computation in the attention read and write kernels; it is the only place
placement enters. Cross-request sharing builds on a content-addressed block
pool with chained hashes verified by exact token comparison, reference
counts, and an LRU free list with revival. Hash collisions can never alias,
because every hit is confirmed by \code{memcmp} over the block's tokens. The
sharing discipline is deliberately simple: shared blocks are immutable,
complete 32-token blocks, the block being appended to is always private, and
a session publishes its blocks to the pool only when it finishes. No
copy-on-write is needed, and speculative rollback can never land inside a
shared block by construction. Sessions of a pooled model borrow one
model-owned physical slab and carry only a block table and token timeline,
so adopting a cached prefix is a block-table edit. The whole mechanism is
gated bitwise against private-slab sessions on both backends
(\Cref{sec:eval-prefix}).

\subsection{Kernels}
\label{sec:impl-kernels}
The Metal backend has three matmul families. Single-token decode uses a
token-tiled simdgroup matvec with one simdgroup per output row and coalesced
loads. Batched prefill uses simdgroup-matrix (MMA) kernels. The BF16 kernel
stages activations as BF16 and walks 32-row by 16-token output tiles with
FP32 accumulation. The quantized MMA kernels dequantize each 32-weight block
into a threadgroup tile once per 16 tokens and run the same fragment loop,
reading activations directly in FP32, so the dequantization cost no longer
scales with token count and quantized prefill tracks BF16 closely
(\Cref{sec:eval-throughput}). Attention comes in two shapes, both in the
FlashAttention lineage
\citep{dao2022flashattention,dao2023flashdecoding}. Decode uses a split-K
streaming online-softmax kernel: eight simdgroups per (head, token) each
scan every eighth source position through the block table, and partial
softmaxes merge in threadgroup memory; no score matrix is materialized.
Past 1{,}024 cached positions the kernel additionally splits each head's
KV scan across up to eight threadgroups, each emitting an online-softmax
partial that a small reduction kernel merges in fixed order; the split
count derives only from a token's own position, so plain decode and
multi-row speculative verify partition identically and short contexts
execute the unsplit path bit-for-bit.
Prefill uses a tiled simdgroup-matrix flash kernel whose 32-position KV
tiles coincide with cache blocks; its arrival also produced the
silent-fallback episode of \Cref{sec:eval-ablation}.

The CUDA backend is smaller, and we state its scope exactly: contiguous KV,
BF16 weights only, final-row logits only, materialized-score attention,
correctness-gated against the oracle on ephemeral cloud GPUs
(\Cref{tab:correct}).
CUDA paging is future work. No Metal code bends to accommodate CUDA or vice
versa.

\subsection{Quantization}
\label{sec:impl-quant}
The seven per-layer projections are stored as either Q8\_0 (8-bit, 34-byte
blocks: fp16 scale, 32 int8 values) or a 4-bit affine scheme (20-byte groups:
16 packed nibbles, fp16 scale, fp16 bias; $w=\text{scale}\cdot q+\text{bias}$)
\citep{frantar2023gptq,lin2024awq,llamacpp2023gguf}. We call the latter
scheme \emph{affine4} (4-bit affine, group size 32) throughout. Embeddings,
the LM head, and norms stay BF16. Both schemes are bit-accurate between the
CPU and Metal implementations, and quantized artifacts pass the same gate
lattice as BF16, including the scramble gates.

\subsection{Speculative decoding with adaptive gating}
\label{sec:impl-spec}
A draft-model-free prompt-lookup matcher \citep{saxena2023promptlookup}
proposes the continuation from the longest recent $n$-gram match. A single
multi-logit forward verifies the whole draft, the longest greedy-matching
prefix is accepted, and rejected drafts are dropped by truncating the KV
back. This is the paged, zero-copy rollback that placement invariance makes
safe. Greedy speculative output is token-identical to non-speculative greedy
by construction, and a gate asserts it.

On a bandwidth-bound engine, speculation at low acceptance is a net loss
(\Cref{sec:bg}): ungated prompt-lookup slows grounded decoding to
$\SpecSpeedupGrounded\times$ (\Cref{tab:spec}). Kipp therefore gates drafting
adaptively (\Cref{alg:gate}). An exponential moving average of per-verify
acceptance suspends drafting when it falls below a threshold, and a periodic
probe re-enables it when the text becomes repetitive again. Adapting
speculation effort to observed acceptance follows SpecDec++
\citep{huang2025specdecpp}, which tunes the draft length online, and BiLD
\citep{kim2023bild}, which uses threshold policies between models; Kipp's
policy is simpler in that it stops drafting entirely and probes. Gating only
chooses between ``draft and verify'' and ``plain decode step.'' Both emit
the greedy sequence, so the token-identity invariant is untouched.

\begin{algorithm}[t]
\caption{Adaptive speculation gating (per decode step)}
\label{alg:gate}
\begin{algorithmic}[1]
\State $\textit{ema} \gets 0.5$;\ \ $\textit{active} \gets \textbf{true}$;\ \
       $\textit{idle} \gets 0$
\For{each decode step}
  \If{$\textit{active}$ \textbf{or} $\textit{idle} \ge P$}
     \Comment{$P{=}32$ tokens between probes}
    \State draft $k$ tokens; verify; accept prefix $a$
    \State $\textit{ema} \gets \tfrac{1}{2}\textit{ema} + \tfrac{1}{2}(a/k)$;
           \ \ $\textit{idle} \gets 0$
    \If{probing \textbf{and} $a/k \ge 0.5$}
       \State $\textit{active} \gets \textbf{true}$;\ \
              $\textit{ema} \gets a/k$
    \ElsIf{$\textit{ema} < 0.2$}
       \State $\textit{active} \gets \textbf{false}$
    \EndIf
  \Else
    \State plain decode step;\ \ $\textit{idle} \gets \textit{idle} + 1$
  \EndIf
\EndFor
\end{algorithmic}
\end{algorithm}

\subsection{Serving}
\label{sec:impl-serving}
A single-threaded \code{poll()} event loop implements the OpenAI Completions
and Chat Completions APIs with continuous batching \citep{yu2022orca},
chunked prefill \citep{agrawal2024sarathi}, a native Qwen3 ChatML renderer,
the full sampling surface, generated-token log-probabilities, and Prometheus
metrics. Multiple choices and concurrent requests decode through one batched
eval per step, so each step reads the weights once for all active sequences;
the admission cap equals the backend batch limit (32 concurrent
generations), so single-choice traffic can fill the whole batch.
The scheduler admits requests FIFO with capacity-based backpressure, holds
streamed bytes back while they could still grow into a stop string, reaps
idle client connections on a deadline, and exports TTFT, queue-wait, and
decode-time counters alongside the pool gauges. The same engine also backs
a CLI with one-shot generation, perplexity scoring, and a multi-turn chat
REPL that re-evaluates only each turn's rendered suffix against the
retained KV.

\section{Evaluation}
\label{sec:eval}

\subsection{Setup and methodology}
\label{sec:eval-setup}
All experiments run on an Apple M5~Max (40-core GPU, 128\,GB unified memory)
with Qwen3-4B and greedy decoding unless stated, using one discarded warm-up
and five measured subprocess runs per configuration. We report medians with
median absolute deviation (MAD). The harness (\code{bench/}) captures
separate prefill and decode timers and peak resident memory, and records the
engine commit, hardware, and model revision in every result file. Earlier
Kipp releases were benchmarked on a base M5 (10-core GPU); all numbers in
this paper were re-measured on the M5~Max, and the committed result files
self-describe their hardware.

\paragraph{Measurement protocol}
Apple-silicon GPU clocks are demand-scaled. Short bursts launched against an
idle GPU measure a fraction of sustained throughput --- we observed
several-fold gaps and large run-to-run swings before controlling for it
% HAND: DVFS calibration measured 2-5x below sustained with ~3x run-to-run
% HAND: swings; protocol details in bench/README.md.
--- so a burst benchmark on a consumer SoC can silently report the ramp, not
the engine. Every number in this paper is sustained steady-state: idle
machine, multi-minute warm-up, back-to-back single-session measurement, and
a MAD-based acceptance bound, as documented in \code{bench/README.md}. The
harness also refuses to record Metal results while the bridge reports a
kernel-compile fallback, a tripwire earned the hard way
(\Cref{sec:eval-ablation}).

For serving experiments we use the standard load metrics
\citep{zhong2024distserve}: time to first token (TTFT), time per output
token (TPOT), and goodput, the fraction of requests meeting both a TTFT and
a TPOT service-level objective. For quantization quality we report WikiText-2
perplexity \citep{li2024evalquant}, computed over non-overlapping
2{,}048-token windows with every position after each window's first scored
and no burn-in; the protocol is deterministic and identical across schemes,
but its absolute values are not comparable to llama.cpp's
\code{perplexity} tool, which discards a per-window burn-in.

\subsection{Correctness gates}
\label{sec:eval-correct}
\Cref{tab:correct} summarizes the gates. The CPU oracle matches the pinned
FP32 reference vectors at exact arg\,max and NMSE $\le 5\times10^{-5}$,
and is byte-stable across engine versions
(\Cref{sec:oracle}). Metal matches the oracle at exact arg\,max and
NMSE near $\MetalNmse$, and paged KV is bitwise identical to a
contiguous cache under an adversarially scrambled block table on both
backends, for all three weight schemes.

\begin{table}[t]\centering\small
\caption{Correctness gates (Qwen3-4B). ``bitwise'' = \code{memcmp}-equal.}
\label{tab:correct}
\begin{tabular}{ll}
\toprule
Gate & Result \\
\midrule
CPU vs.\ pinned FP32 reference   & arg\,max exact, NMSE ${\le}5{\cdot}10^{-5}$ \\
CPU vs.\ its own prior releases  & bitwise identical \\
Metal vs.\ CPU oracle            & arg\,max exact, NMSE $\MetalNmse$ \\
CUDA vs.\ CPU oracle (H100)$^a$  & arg\,max exact, NMSE ${\le}\CudaMaxNmse$ \\
Paged vs.\ contiguous (CPU)      & bitwise, scrambled table \\
Paged vs.\ contiguous (Metal)    & bitwise, scrambled table \\
Spec.\ vs.\ greedy decode        & token-identical \\
\bottomrule
\multicolumn{2}{l}{\scriptsize $^a$ CUDA scope: contiguous KV, BF16
weights, final-row logits. Revalidated on an} \\
\multicolumn{2}{l}{\scriptsize \ \ ephemeral cloud \CudaGpuName{} across
\CudaCheckpointCount{} checkpoints} \\
\multicolumn{2}{l}{\scriptsize \ \ (\code{bench/results/cuda-h100-gates.json}).}
\end{tabular}
\end{table}

\subsection{Fault injection: do the gates catch bugs?}
\label{sec:eval-faults}
The claim of \Cref{sec:silent} is empirical: tolerance-based reference
comparisons mask realistic paging bugs that the scramble gate catches. We
test it by mutation \citep{demillo1978mutation,jia2011mutation}. A test-only
build seeds one deliberate KV-addressing bug at a time
\citep{hsueh1997fault}, and we run both gate families against each mutant.
The faults are asymmetric, affecting the read side or the write side only,
because \Cref{prop:invariance} itself proves that a mapping error applied
identically to both sides is a harmless relocation. The mutation study must
model where real bugs live, in one side's copy of the indirection:

\begin{enumerate}\itemsep1pt
  \item \emph{read-block}: attention reads resolve the next logical block;
  \item \emph{read-slot}: attention reads resolve the next slot within the
  block;
  \item \emph{rollover}: each block's first position is written one past
  the end of the previous block's physical block, so under the identity
  table those bytes land exactly where the position belongs;
  \item \emph{swap-kv}: one layer writes keys into the value store and
  vice versa.
\end{enumerate}

\Cref{tab:faults} reports, for each mutant, the observed NMSE against the
reference, whether the final arg\,max survived, and each gate's verdict.
The detection pattern partitions the faults along the lines of
\Cref{prop:invariance}.

The placement-commuting faults (read-block, read-slot, swap-kv) corrupt the
same logical positions under every block table, so both scramble runs
compute identical wrong logits and the bitwise gate passes. The proposition
predicts this: the scramble gate cannot see them. Reference testing catches
all three, but in two distinct ways. The read-side faults corrupt the
logits at NMSE near $10^{-2}$ \emph{without} flipping the final arg\,max
--- only the NMSE bound sees them --- so a gate that checked arg\,max
alone, a common lightweight regression test, would have passed every
mutant except swap-kv, whose gross corruption (NMSE
$\FaultNmseSwapKv$, arg\,max flipped) any check catches.

The rollover fault is the mirror image. Its misaddressed write lands on
exactly the right byte under the identity table, so the reference comparison
measures NMSE of exactly $\FaultNmseRollover$, and no tolerance, however
tight, can ever detect it. Only the scramble gate rejects this fault: under
a non-identity table the same bug scatters the write into the wrong physical
block. The two gate families are therefore complementary rather than
redundant. The reference gate covers value corruption, and the scramble gate
covers the placement-coupled residue that reference testing cannot reach.
That residue is precisely the class of bug that ships silently today,
because it only corrupts once a scheduler starts relocating blocks in
production.

The study also caught the gate itself. Our original scramble gate spanned
two blocks, and the rollover mutant passed it: reversing a 2-entry table is
a swap, and ``one past the end of the previous block's physical block,
wrapping'' happens to land on the correct byte again. Reversal over three or
more blocks breaks physical adjacency and exposes the fault, so the gate now
spans three. Mutation testing evaluated the adequacy of the tests as well as
the engine, and improved one of the tests.

\begin{table}[t]\centering\small
\caption{Fault-detection matrix (Qwen3-4B, CPU oracle, 96-token/3-block
sequence). ``caught'' = the gate fails on the mutant. Values from
\code{bench/results/faults.json}.}
\label{tab:faults}
\begin{tabular}{lrccc}
\toprule
Seeded fault & NMSE & arg\,max & Tolerance gate & Scramble gate \\
\midrule
(control)    & $\FaultControlNmse$        & same    & clean   & clean \\
read-block   & $\FaultNmseReadBlock$ & same & caught  & \emph{missed} \\
read-slot    & $\FaultNmseReadSlot$ & same & caught  & \emph{missed} \\
rollover     & $\FaultNmseRollover$        & same    & \emph{missed} & \textbf{caught} \\
swap-kv      & $\FaultNmseSwapKv$      & flipped & caught  & \emph{missed} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Throughput and quality by weight scheme}
\label{sec:eval-throughput}
\Cref{tab:perf} reports decode and prefill throughput by weight scheme.
Decode is bandwidth-bound, so quantization scales it as it shrinks the
weight bytes streamed per token. Prefill is compute-bound, and the
quantized MMA kernels hold it near BF16 (\PrefillQEight{} and
\PrefillAffine{} vs.\ \PrefillBF~tok/s, within $4$--$8\%$): dequantizing
each 32-weight block once per 16-token tile keeps the added arithmetic off
the critical path, so quantization buys its decode speedup without
surrendering prefill.

Speed is only half the tradeoff. \Cref{fig:quality} plots each scheme's
WikiText-2 perplexity against its decode throughput, which is the plot a
deployment decision actually needs: how much quality each streamed byte
buys. Q8\_0 is quality-free at this precision of measurement
(\PplQEight{} vs.\ \PplBF{} perplexity, $+\PplDeltaQEightPct\%$) while
lifting decode from \DecodeBF{} to \DecodeQEight~tok/s; affine4 pays
$+\PplDeltaAffinePct\%$ perplexity (\PplAffine) for the fastest decode of
the three (\DecodeAffineFour~tok/s).

\begin{table}[t]\centering\small
\caption{Throughput (tok/s), Qwen3-4B, Apple M5~Max Metal, median of 5.
Median absolute deviations are ${\le}1$ tok/s throughout (per-run
distributions in the committed result files).}
\label{tab:perf}
\begin{tabular}{lrr}
\toprule
Weight scheme & Decode & Prefill \\
\midrule
BF16     & \DecodeBF         & \PrefillBF \\
Q8\_0    & \DecodeQEight     & \PrefillQEight \\
affine4  & \DecodeAffineFour & \PrefillAffine \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[t]\centering
\begin{tikzpicture}
\begin{axis}[kippplot,
  xlabel={decode throughput (tok/s)},
  ylabel={WikiText-2 perplexity},
  xmin=40, xmax=150, ymin=7.5, ymax=8.5,
]
\addplot+[only marks, mark size=2.4pt]
  table[x=decode_tps, y=ppl] {data/quant-ppl.dat};
\node[font=\scriptsize, anchor=south, yshift=2pt]
  at (axis cs:\DecodeBF,\PplBF) {BF16};
\node[font=\scriptsize, anchor=south, yshift=2pt]
  at (axis cs:\DecodeQEight,\PplQEight) {Q8\_0};
\node[font=\scriptsize, anchor=south, yshift=2pt]
  at (axis cs:\DecodeAffineFour,\PplAffine) {affine4};
\end{axis}
\end{tikzpicture}
\caption{Quality--speed tradeoff by weight scheme (Qwen3-4B, M5~Max Metal):
WikiText-2 perplexity against decode throughput, one point per scheme.
Down and to the right is better.}
\Description{Scatter plot of perplexity versus decode throughput for three
weight schemes.}
\label{fig:quality}
\end{figure}

\subsection{Model-size scaling}
\label{sec:eval-modelsize}
Decode throughput on bandwidth-bound hardware should scale inversely with
the bytes streamed per token, independent of which checkpoint streams them.
\Cref{fig:modelsize} checks this across the family (0.6B, 4B, 8B), so the
4B results of \Cref{sec:eval-throughput} can be read as representative
rather than cherry-picked: BF16 decode falls from \DecodeZeroSixBBF{}
through \DecodeBF{} to \DecodeEightBBF~tok/s as parameters grow
$0.6\to4\to8$B, and Q8\_0 halves the streamed bytes at 8B for
\DecodeEightBQEight~tok/s.

\begin{figure}[t]\centering
\begin{tikzpicture}
\begin{axis}[kippplot,
  xmode=log, ymode=log, log ticks with fixed point,
  xlabel={parameters (billions)}, ylabel={tok/s},
  xmin=0.4, xmax=12,
  legend columns=2,
  legend style={at={(0.5,1.02)}, anchor=south},
]
% Every coordinate below is a generated macro bound to model-size.dat's
% source JSONs (06b-*, 4b-*, 8b-* decode/prefill).
\addplot coordinates {(0.6,\DecodeZeroSixBBF) (4,\DecodeBF)
  (8,\DecodeEightBBF)};
\addplot coordinates {(0.6,\PrefillZeroSixBBF) (4,\PrefillBF)
  (8,\PrefillEightBBF)};
\addplot coordinates {(4,\DecodeQEight) (8,\DecodeEightBQEight)};
\addplot coordinates {(4,\PrefillQEight) (8,\PrefillEightBQEight)};
\legend{decode BF16, prefill BF16, decode Q8\_0, prefill Q8\_0}
\end{axis}
\end{tikzpicture}
\caption{Throughput across the Qwen3 dense family (M5~Max Metal). Decode
tracks streamed weight bytes per token; prefill tracks compute.}
\Description{Line plot of decode and prefill throughput against model
parameter count.}
\label{fig:modelsize}
\end{figure}

\subsection{Context scaling}
\label{sec:eval-context}
\Cref{fig:ctx} reports throughput against prompt length (Q8\_0, 32-token
decode, median of 3; \code{bench/results/ctx-*.json}). Decode degrades
gracefully, from \CtxDecodeShort{} to \CtxDecodeLongest~tok/s between a 3-
and a 12{,}800-token context, as the per-step KV read grows against the
fixed weight stream; the KV-length split of \Cref{sec:impl-kernels} is
what holds the 12{,}800-token point near two-thirds of the bandwidth
roofline, where a single-threadgroup-per-head scan reaches about a third. Prefill peaks near \CtxPrefillPeak~tok/s at an
800-token prompt and then declines steadily with context, the expected
shape once attention's quadratic term grows against the linear matmul
work; at 12{,}800 tokens prefill still sustains over a third of its peak.

\begin{figure}[t]\centering
\begin{tikzpicture}
\begin{axis}[kippplot, height=4.8cm,
  xmode=log, log ticks with fixed point,
  xlabel={prompt tokens}, ylabel={tok/s}, ymin=0,
  legend columns=2,
  legend style={at={(0.5,1.02)}, anchor=south},
]
\addplot+[error bars/.cd, y dir=both, y explicit]
  table[x=tokens, y=decode, y error=decode_mad] {data/ctx-scaling.dat};
\addplot+[error bars/.cd, y dir=both, y explicit]
  table[x=tokens, y=prefill, y error=prefill_mad] {data/ctx-scaling.dat};
\legend{decode, prefill}
\end{axis}
\end{tikzpicture}
\caption{Context scaling (Qwen3-4B Q8\_0, M5~Max Metal; error bars show
MAD). Decode falls as the per-step KV read grows; prefill peaks at short
context and declines as attention's quadratic cost grows against the
linear matmul work.}
\Description{Line plot of decode and prefill throughput against prompt
length on a log axis.}
\label{fig:ctx}
\end{figure}

\subsection{Batch scaling}
\label{sec:eval-batching}
Because decode is bandwidth-bound, aggregate server throughput scales with
the number of concurrently decoding sequences: each batched step reads the
weights once for all choices. \Cref{fig:batch} plots aggregate throughput on
the pooled Q8\_0 server for one request with $n$ parallel choices and for
$n$ independent concurrent connections, against the ideal of $n$ times the
single-stream rate (sampled decoding, median of 3;
\code{bench/results/server-batch.json}). One request with $n{=}1/2/4/8$ choices delivers
\BatchChoicesTpsOne, \BatchChoicesTpsTwo, \BatchChoicesTpsFour, and
\BatchChoicesTpsEight~tok/s aggregate; $1/2/4/8$ independent connections
deliver \BatchConnsTpsOne, \BatchConnsTpsTwo, \BatchConnsTpsFour, and
\BatchConnsTpsEight.
The two series differ at $n{=}1$ because they measure different
aggregates: parallel choices share one prompt's prefill across the batch,
while concurrent connections each pay their own prefill inside the
measured window (the ideal-linear reference is therefore anchored on the
choices series).
The shapes then diverge instructively. Independent connections scale
monotonically through $n{=}8$, as batched decode amortizes the weight
stream across every admitted request. Parallel
choices peak at $n{=}2$ and then \emph{decline}: each additional choice
adds per-step CPU-side sampling work in the single-threaded scheduler
(these runs sample at temperature $0.8$, so the greedy fast path does not
apply), the GPU idles longer between batched steps, and the demand-scaled
clocks of \Cref{sec:eval-setup} sag in response --- a
serving-on-consumer-SoC effect a fixed-clock datacenter GPU would not show.
The gap to the ideal line is that scheduling cost made visible.

\begin{figure}[t]\centering
\begin{tikzpicture}
\begin{axis}[kippplot,
  xlabel={concurrently decoding sequences $n$},
  ylabel={aggregate tok/s},
  xtick={1,2,4,8}, xmin=0.6, xmax=8.4, ymin=0, ymax=200,
  legend columns=1,
  legend style={at={(0.97,0.03)}, anchor=south east},
]
\addplot table[x=n, y=choices_tps] {data/server-batch.dat};
\addplot table[x=n, y=connections_tps] {data/server-batch.dat};
\addplot[dashed, gray, domain=1:2.55, samples=2] {\BatchChoicesTpsOne*x};
\legend{parallel choices, concurrent connections, ideal linear}
\end{axis}
\end{tikzpicture}
\caption{Batch scaling on the pooled Q8\_0 server (M5~Max). The dashed line
is ideal linear scaling from the single-stream rate. Independent
connections scale monotonically; parallel choices peak at $n{=}2$, where
per-step CPU-side sampling begins stalling the GPU long enough for
demand-scaled clocks to sag.}
\Description{Line plot of aggregate throughput against number of concurrent
sequences, with a dashed ideal-scaling reference.}
\label{fig:batch}
\end{figure}

\subsection{Serving under load}
\label{sec:eval-load}
Batch scaling measures a closed system; serving is an open one. We drive the
server with open-loop Poisson arrivals of streamed completion requests at
increasing offered rates and record per-request TTFT and TPOT percentiles
and goodput under a fixed SLO (\Cref{fig:load};
\code{bench/results/load.json}). This is the view a capacity planner needs:
the offered rate at which tail latency leaves the knee, and how much of the
offered load still meets its SLO beyond it.

The verdict for this workload (32-request bursts of 348-token prompts): at
$0.5$~req/s the server holds its service-level objective for the large
majority of requests (median TTFT \LoadTtftPFiftyLow~ms), and goodput
collapses to \LoadGoodputHigh{} from $2$~req/s on. Past the knee the
overload lands where an admission-controlled FIFO puts it: p99 TTFT grows
monotonically with queue depth to \LoadTtftPNinetyNineHigh~s at
\LoadRateHigh~req/s, aggregate output tops out near \LoadPeakTps~tok/s,
and per-token latency rises only with batch occupancy --- TPOT roughly
triples from the lowest to the highest offered rate as more requests share
each batched decode step, then flattens (\Cref{fig:load}, right). The
system behaves exactly as the
batch-scaling analysis predicts --- admission control and the FIFO queue
protect correctness, decode throughput degrades only by batch sharing, and
time to first token absorbs the overload.

\begin{figure*}[t]\centering
\begin{tikzpicture}
\begin{axis}[kippplot, width=0.44\textwidth,
  xlabel={offered rate (req/s)}, ylabel={TTFT (s)},
  xmin=0.4, xmax=4.2, ymin=0, ymax=20,
  legend columns=3,
  legend style={at={(0.97,0.03)}, anchor=south east},
]
\addplot table[x=rate, y=ttft_p50] {data/load-latency.dat};
\addplot table[x=rate, y=ttft_p90] {data/load-latency.dat};
\addplot table[x=rate, y=ttft_p99] {data/load-latency.dat};
\legend{p50, p90, p99}
\end{axis}
\end{tikzpicture}\hfill
\begin{tikzpicture}
\begin{axis}[kippplot, width=0.44\textwidth,
  xlabel={offered rate (req/s)}, ylabel={TPOT (ms)},
  xmin=0.4, xmax=4.2, ymin=0, ymax=500,
  legend columns=2,
  legend style={at={(0.97,0.03)}, anchor=south east},
]
\addplot table[x=rate, y expr=\thisrow{tpot_p50}*1000] {data/load-latency.dat};
\addplot table[x=rate, y expr=\thisrow{tpot_p99}*1000] {data/load-latency.dat};
\legend{p50, p99}
\end{axis}
\end{tikzpicture}
\caption{Serving under open-loop Poisson load (Qwen3-4B Q8\_0, pooled
server, M5~Max): TTFT and TPOT percentiles against offered request rate.
TTFT grows with queue depth; TPOT rises with batch occupancy and then
flattens --- the overloaded server queues and shares each decode step, it
does not thrash.}
\Description{Two panels of latency percentiles against offered request
rate.}
\label{fig:load}
\end{figure*}

\subsection{Cross-request prefix sharing}
\label{sec:eval-prefix}
Prefix sharing is live in the serving path. A finished session publishes its
full 32-token blocks to the content-addressed pool, and a new request adopts
the longest published prefix at admission, re-evaluating only the partial
tail. On a \PrefixPromptTokens-token prompt sent twice
(\code{bench/results/prefix-reuse.json}), the second request adopts
\PrefixAdoptedTokens{} tokens and its prefill time collapses from
\PrefixColdSeconds~s to \PrefixWarmMs~ms, a $\PrefixSpeedup\times$ reduction
in time to first token, with byte-identical output.

The correctness side is gated rather than assumed: \Cref{tab:pooled} lists
the pooled-gate scenarios, each proven on both backends for BF16 and
quantized weights. This is the payoff of \Cref{sec:vpa}: block adoption is a
block-table edit, and the scramble gate is what makes editing tables a safe
operation.

\begin{table}[t]\centering\small
\caption{Pooled-sharing gate scenarios (\code{--pooled-cpu},
\code{--pooled-metal}), each run on CPU and Metal for BF16, Q8\_0, and
affine4 weights.}
\label{tab:pooled}
\begin{tabular}{@{}lp{3.6cm}@{}}
\toprule
Scenario & Criterion \\
\midrule
Serial prefix sharing & bitwise vs.\ private-slab session \\
Mixed shared/private batch & bitwise vs.\ private-slab sessions \\
Pool exhaustion & clean failure; admission delays, never corrupts \\
Truncation (spec.\ rollback) & bitwise; never enters a shared block \\
Eviction and revival & bitwise after LRU eviction \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Speculative decoding}
\label{sec:eval-spec}
\Cref{tab:spec} and \Cref{fig:spec} report acceptance rate $\alpha$, block
efficiency (accepted tokens per target forward), and wall-clock speedup
across four workloads, with and without adaptive gating. The A/B is
paired: for each workload the harness measures plain decode, ungated
speculation, and gated speculation adjacently within one session, so both
speedup columns share a single baseline (the gap between the two
mechanically identical repetitive configurations, $\SpecSpeedupRep\times$
vs.\ $\GatedSpeedupRep\times$ at duty $1.0$, bounds residual run-to-run
variance). One honest denominator note: a greedy-sampling fast path landed
in the same revision as this measurement session, so the plain-decode
baselines are faster than in earlier drafts and every speculation ratio is
correspondingly compressed --- same absolute wall-clock wins, smaller
multipliers. Ungated prompt-lookup lifts
copy-heavy decode from \SpecBaseRep{} to \SpecDecodeRep~tok/s
($\SpecSpeedupRep\times$) at $\alpha{=}1.0$. However, it slows code to
$\SpecSpeedupCode\times$ and grounded text to as little as
$\SpecSpeedupGrounded\times$: the $k{+}1$-token verify does not amortize
weight streaming at small draft batch, so each speculative step costs close
to $k{+}1$ decodes while advancing by only accepted${}+1$ tokens. The
adaptive gate suspends drafting on exactly those workloads --- its duty
cycle (the fraction of steps that draft) makes the mechanism visible ---
and holds every workload at or above $\GatedSpeedupChat\times$, pushing
code past parity ($\GatedSpeedupCode\times$) while keeping the copy-heavy
win ($\GatedSpeedupRep\times$). In every configuration the emitted tokens
are identical to greedy decoding; speculation changes only the wall clock.

\begin{table}[t]\centering\small
\caption{Prompt-lookup speculation by workload (Qwen3-4B Q8\_0, M5~Max,
256-token decode), controlled A/B, median of 3. ``Gated'' adds the adaptive
gate of \Cref{alg:gate}; ``duty'' is its drafting duty cycle. Speedup ${<}1$
is slower than plain decode.}
\label{tab:spec}
\begin{tabular}{lrrrrr}
\toprule
Workload & $\alpha$ & Blk.\ eff. & Ungated & Gated & Duty \\
\midrule
Repetitive & \SpecAlphaRep & \SpecBlkRep &
  $\SpecSpeedupRep\times$ & $\GatedSpeedupRep\times$ & \GatedDutyRep \\
Code       & \SpecAlphaCode & \SpecBlkCode &
  $\SpecSpeedupCode\times$ & $\GatedSpeedupCode\times$ & \GatedDutyCode \\
Grounded   & \SpecAlphaGrounded & \SpecBlkGrounded &
  $\SpecSpeedupGrounded\times$ & $\GatedSpeedupGrounded\times$ & \GatedDutyGrounded \\
Open-ended & \SpecAlphaChat & \SpecBlkChat &
  $\SpecSpeedupChat\times$ & $\GatedSpeedupChat\times$ & \GatedDutyChat \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[t]\centering
\begin{tikzpicture}
\begin{axis}[kippplot,
  ybar, bar width=6pt, ymin=0,
  ylabel={speedup vs.\ plain decode},
  symbolic x coords={repetitive,code,grounded,chat},
  xtick=data, enlarge x limits=0.15,
  legend columns=2,
  legend style={at={(0.5,1.02)}, anchor=south},
]
\addplot[ybar, fill=kippblue!75, draw=kippblue]
  table[x=workload, y=ungated] {data/spec-speedup.dat};
\addplot[ybar, fill=kipporange!75, draw=kipporange]
  table[x=workload, y=gated] {data/spec-speedup.dat};
\draw[dashed, gray]
  ({rel axis cs:0,0}|-{axis cs:repetitive,1}) --
  ({rel axis cs:1,0}|-{axis cs:repetitive,1})
  node[pos=0.88, above, font=\tiny, text=gray] {parity};
\legend{ungated, gated}
\end{axis}
\end{tikzpicture}
\caption{Speculation speedup by workload, with and without the adaptive
gate. The dashed line is parity with plain decode: gating preserves the
repetitive-workload win, pushes code past parity, and pulls the remaining
workloads back to it.}
\Description{Grouped bar chart of speculation speedup per workload, ungated
and gated, with a parity line at one.}
\label{fig:spec}
\end{figure}

\subsection{Comparison with llama.cpp}
\label{sec:eval-llamacpp}
We compare against llama.cpp \citep{llamacpp2023gguf} on identical hardware
and the same pinned Qwen3-4B revision, converted independently with
llama.cpp's own tooling, at the pinned llama.cpp commit recorded with the
exact commands in \code{bench/results/llamacpp-qwen3-4b.json}
(\code{llama-bench}, 2{,}048-token prefill, 256-token decode, five
repetitions). \Cref{tab:llamacpp} pairs llama.cpp's numbers with Kipp's
decode benchmark and Kipp's matched 2{,}048-token prefill. The result cuts
both ways. On decode, the phase that dominates serving cost, Kipp is faster
in every matched scheme, about $\LlamaDecodeRatioQEight\times$ at 8-bit
(\DecodeQEight{} vs.\ \LlamaDecodeQEight~tok/s), and its affine4 outruns
llama.cpp's Q4\_0 (\DecodeAffineFour{} vs.\ \LlamaDecodeQFour). The
verification discipline costs nothing where it matters most. On prefill
llama.cpp leads, about $\LlamaPrefillRatioBF\times$ at
BF16 (\LlamaPrefillBF{} vs.\ \PrefillTwoKBF~tok/s at 2{,}048 tokens) and
somewhat more at the quantized schemes, reflecting years of kernel
engineering (larger tiles, threadgroup-level pipelining, autotuned
dispatch) that
Kipp's readable single-kernel-per-op design does not attempt. Kipp is not
the fastest prefill engine and does
not try to be; its scope is one pinned family with no autotuning, and we
present this number as the measured price of a readable, verifiable prefill
path today. Three caveats on fairness. Both engines were measured in the
same idle steady-state session (\Cref{sec:eval-setup}), but
\code{llama-bench} reports
mean$\pm$stddev while Kipp reports median$\pm$MAD. The
decode configurations differ (64- vs.\ 256-token generations), but
batch-one decode rate is insensitive at these depths --- our context sweep
moves only from \CtxDecodeShort{} to \CtxDecodeTwoHundred~tok/s between 3
and 200 tokens of context --- so both measure the same steady rate. And the 4-bit schemes
differ (llama.cpp Q4\_0 is scale-only; affine4
carries a bias), so that column is close cousins rather than twins.

\begin{table}[t]\centering\small
\caption{llama.cpp head-to-head (Qwen3-4B, M5~Max Metal; tok/s). Decode:
Kipp median of 5 (348-token prompt, 64-token decode) vs.\
\code{llama-bench} mean of 5 ($n{=}256$). Prefill: matched 2{,}048-token
prompt on both. llama.cpp Q4\_0 is compared against Kipp affine4 (schemes
differ; see text).}
\label{tab:llamacpp}
\begin{tabular}{lrrrr}
\toprule
& \multicolumn{2}{c}{Decode} & \multicolumn{2}{c}{Prefill (2k)} \\
\cmidrule(lr){2-3}\cmidrule(lr){4-5}
Weights & Kipp & llama.cpp & Kipp & llama.cpp \\
\midrule
BF16    & \textbf{\DecodeBF} & \LlamaDecodeBF & \PrefillTwoKBF & \textbf{\LlamaPrefillBF} \\
Q8\_0   & \textbf{\DecodeQEight} & \LlamaDecodeQEight & \PrefillTwoKQEight & \textbf{\LlamaPrefillQEight} \\
affine4 / Q4\_0 & \textbf{\DecodeAffineFour} & \LlamaDecodeQFour & \PrefillTwoKAffine & \textbf{\LlamaPrefillQFour} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Memory}
\label{sec:eval-memory}
Peak process RSS is about \PeakRSS~MB for the 4B model regardless of
quantization, since the loader \code{mmap}s weights and pages them in on
demand. Note that peak RSS understates true unified-memory residency for
GPU-touched \code{mmap} pages, so we report artifact size alongside RSS and
never present RSS as the total memory needed to run the model.

\subsection{Ablations}
\label{sec:eval-ablation}
\paragraph{A silent fallback defeats every correctness gate} The sharpest
lesson in this paper was self-inflicted, and we report it in full. The
tiled flash-prefill kernel of \Cref{sec:impl-kernels} landed with a loop
variable named \code{fragment} --- a reserved identifier in the Metal
Shading Language. The runtime compile of the \emph{entire} simdgroup-matrix
library therefore failed with a syntax error, and the bridge silently fell
back to vector kernels for every matrix matmul \emph{and} for the new
kernel itself, which as a result had never actually executed: its gates,
and an entire benchmark campaign, exercised the fallback. The fallback is a
correct implementation, so every correctness gate in the lattice passed
throughout --- while prefill shipped 35--75\% slower than the engine's real
capability, quantized prefill appeared to trail BF16 at exactly the
superseded vector path's throughput, and context-scaling appeared to
collapse. The only symptom was one warning line on \code{stderr}, which
every harness discarded. A one-identifier rename restored everything: BF16
prefill at a 348-token prompt rose from 355 to \PrefillBF~tok/s, Q8\_0 from
218 to \PrefillQEight{} (restoring the quantized-MMA near-parity the design
had always provided), and Q8\_0 at 2{,}048 tokens from 70 to
\PrefillTwoKQEight.
% HAND: the broken-build values (355 / 218 / 70) are historical, measured
% HAND: on the fallback build before commit a102e1c; that campaign's
% HAND: result files were superseded and are not in the committed set.
The moral sharpens this paper's thesis rather than softening it:
correctness gating is necessary but not sufficient. A silent within-engine
fallback passes every value-level check by construction, so fallbacks must
be loud, and performance needs its own tripwire --- the harness now refuses
to record Metal results while the fallback warning is present. Every kernel
change in this line was, and remains, admitted through the same lattice as
any other feature: exact arg\,max, per-scheme NMSE, bitwise scramble.
\paragraph{Verify-path reduction order} The MMA work also produced a
cautionary result. Routing speculative verify rounds through the matrix
kernels broke the spec-equals-greedy gate, because the different summation
order flipped near-tie arg\,maxes. Verify rounds now force the vector
kernels, whose per-token reduction order is bitwise-identical to decode's,
and the gate is green again. The contract caught its own optimization.
\paragraph{Gate sensitivity} The mutation study of \Cref{sec:eval-faults}
doubles as an ablation of the gate lattice. Removing the scramble gate
leaves the rollover fault undetectable in principle. Removing the NMSE bound
and keeping only arg\,max would pass three of the four mutants. Shrinking
the scramble sequence back to two blocks re-masks the rollover fault. Every
component of the lattice has at least one mutant that only it catches.
\paragraph{Adaptive gating} The paired columns of \Cref{tab:spec} isolate
the gate's effect: it preserves the repetitive-workload win, pushes code
past parity, and converts the remaining low-acceptance losses to
near-parity.

\section{Discussion and Limitations}
\label{sec:discussion}

\paragraph{What the contract does not cover} Placement invariance is
claimed only for retained KV, not for logical reordering or prefill-column
permutation (\Cref{sec:vpa}). The scramble gate exercises one adversarial
permutation (reversal) per run. It localizes one-sided addressing bugs, but
a bug that only manifests under a specific table shape could in principle
require other permutations. These are cheap to add, since any non-identity
permutation is a valid gate.

\paragraph{Porting the discipline, and future work} Nothing in the scramble
gate is specific to Kipp. Any paged engine with a per-sequence block table
can evaluate a multi-block prompt twice, once under identity and once under
a reversal, and \code{memcmp} the logits, provided its kernels satisfy H3
within a run (fixed reduction order for a fixed batch shape), which
batch-invariant kernels \citep{he2025defeating} already provide. The
silent-fallback episode of \Cref{sec:eval-ablation} argues for a companion
discipline: any within-engine fallback path should fail loudly or be
recorded in measurement provenance, because value-level gates cannot see
it. We would
welcome the gate's adoption in vLLM or SGLang; it is a few dozen lines
against either engine's test harness. The most promising extension is
tree-shaped speculative rollback
\citep{cai2024medusa,li2024eagle,li2024eagle2}: chain methods roll back by
cheap truncation and dense-buffer tree methods pay a gather-copy, so
reclaiming rejected tree branches by block-table unlinking, under a
bitwise-safety gate, is exactly the kind of operation placement invariance
makes trustworthy.

\subsection{Limitations}
\label{sec:limitations}
\begin{itemize}\itemsep2pt
  \item \emph{Scope.} Kipp is a single-node, single-GPU engine for one
  pinned model family; distributed serving and heterogeneous placement are
  out of scope.
  \item \emph{CUDA.} The CUDA backend is contiguous, BF16-only, final-row
  logits, gated on ephemeral cloud GPUs (most recently a \CudaGpuName{}
  across \CudaCheckpointCount{} checkpoints, \Cref{tab:correct}).
  Extending paging and the scramble gate to CUDA
  is mechanical (the kernels mirror Metal's structure) but unverified, so we
  do not claim it.
  \item \emph{One permutation per run.} The scramble gate reverses the
  table; other adversarial permutations are possible and cheap but not yet
  exercised.
  \item \emph{Quality metric.} Quantization quality is evaluated by
  perplexity only; task-suite accuracy \citep{li2024evalquant} is future
  work.
  \item \emph{Oracle scope.} The oracle pins greedy logits; it does not
  characterize sampling distributions beyond the logits it checks.
\end{itemize}

\section{Related Work}
\label{sec:related}
PagedAttention \citep{kwon2023pagedattention} introduced the block table,
and SGLang's RadixAttention \citep{zheng2024sglang} shares prefixes across
calls via a radix tree. A growing family of systems moves KV around at
runtime: Llumnix migrates running requests and their caches between
instances \citep{sun2024llumnix}, InfiniGen spills KV between GPU and host
memory \citep{lee2024infinigen}, Mooncake ships it between prefill and
decode clusters \citep{qin2025mooncake}, CachedAttention persists it across
conversation turns \citep{gao2024cachedattention}, and Prompt Cache reuses
precomputed attention states across prompts \citep{gim2024promptcache}.
vAttention revisits the layout question entirely, backing contiguous virtual
KV with demand paging \citep{prabhu2025vattention}, and production engines
such as TensorRT-LLM \citep{nvidia2023tensorrtllm} and TGI
\citep{huggingface2023tgi} relocate blocks internally. All of these systems
evaluate throughput and cache-hit rate; none tests placement invariance as a
named property (\Cref{tab:survey}). CacheGen \citep{liu2024cachegen}
makes an interesting contrast, since it compresses KV lossily in transit and
deliberately gives up the bitwise equality that our gate demands. On-device
engines such as llama.cpp \citep{llamacpp2023gguf} and MLX
\citep{hannun2023mlx} target the same hardware class as Kipp's Metal
backend.

\citet{he2025defeating} trace run-to-run nondeterminism to
batch-size-dependent kernel reductions and restore bitwise determinism with
batch-invariant kernels. That work fixes the reduction-order axis; our
contract fixes the orthogonal placement axis and adds a byte-stable scalar
oracle as a hard cross-backend gate. On the testing side, the scramble gate
is a metamorphic test \citep{chen2018metamorphic}, and the oracle discipline
is differential testing \citep{mckeeman1998differential} in the lineage of
CRADLE's cross-framework validation \citep{pham2019cradle}. Symbolic
checkers such as GKLEE \citep{li2012gklee} verify GPU kernels exhaustively
but do not scale to a full transformer forward pass, and fuzzers for
deep-learning libraries \citep{wei2022freefuzz} find crashes rather than
wrong-computation bugs, which require an oracle. Our fault study applies
classical mutation testing \citep{demillo1978mutation,jia2011mutation} and
software fault injection \citep{hsueh1997fault} to an inference engine's
memory system. MLPerf \citep{reddi2020mlperf} standardizes performance
reporting; we extend the same rigor to correctness claims.

FlashAttention
\citep{dao2022flashattention,dao2024flashattention2,shah2024flashattention3}
made exact attention IO-efficient; Flash-Decoding
\citep{dao2023flashdecoding} specializes the split-K streaming shape for
decode, and FlashInfer \citep{ye2025flashinfer} generalizes paged and
block-sparse attention kernels behind a serving-oriented interface. Kipp's
streaming online-softmax Metal kernel is in this lineage. Post-training
quantization \citep{frantar2023gptq,lin2024awq} and GGUF k-quants
\citep{llamacpp2023gguf} inform Kipp's block-quantized weights, and
\citet{li2024evalquant} systematize quality evaluation for quantized LLMs.
Speculative decoding
\citep{leviathan2023speculative,chen2023speculativesampling} accelerates
decode, with draft-model-free \citep{saxena2023promptlookup,fu2024lookahead},
self-drafting \citep{cai2024medusa,li2024eagle,li2024eagle2}, and adaptive
\citep{huang2025specdecpp,kim2023bild} variants. Kipp's contribution here is
the acceptance-gated policy and the token-identity gate, not a new drafting
scheme. DistServe \citep{zhong2024distserve}
defines the goodput-oriented serving metrics our load evaluation adopts.

\section{Conclusion}
Kipp shows that the placement invariance every paged inference system relies
on can be a tested contract rather than an assumption, and that the stakes
are measurable by mutation. A realistic block-rollover bug is bitwise
invisible to every identity-mapped reference test, yet a scramble gate that
costs seconds on GPU catches it immediately, while read-side corruption
sails past arg\,max checks and falls only to a tight NMSE bound. The gates
are complementary, each with mutants that only it catches, and the mutation
study repaired a blind spot in the gate itself. A byte-stable scalar oracle
holds real GPU backends to exact arg\,max, and every feature in the engine,
from quantization and MMA prefill to batching, adaptive speculation, and
pooled KV sharing, landed through the same lattice. The result is a small,
continuously verifiable engine whose correctness, and every benchmark
number in this paper, is reproducible by re-execution --- and a discipline
that any paged engine can adopt for the cost of a reversed block table and
a \code{memcmp}.

\section*{Artifact Availability}
Kipp is open source under the MIT license:\par
\noindent\url{https://github.com/Polished-Snow/Kipp}\par
\noindent This paper was built from commit \texttt{553197b} of that
repository, released as tag \texttt{v0.0.3}. A Zenodo archive of that
tagged tree, with a DOI, is the citable snapshot the reproduction
commands below run against (\code{.zenodo.json} and
\code{paper/SUBMITTING.md} describe the deposit).
\code{REPRODUCE.md} in the repository
documents the full build, gate, and benchmark reproduction path, and
\Cref{app:artifact} maps each claim in this paper to the command that checks
it. Every measured number is generated into the paper from a committed
\code{bench/results/*.json} by \code{make paper-data} and verified by
\code{make paper-check}.

\balance
\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\appendix
\section{Artifact: claims to commands}
\label{app:artifact}
Every correctness claim maps to a gate command, and every number maps to a
committed result file. Build and gate commands (see also
\code{REPRODUCE.md}):

\begin{lstlisting}
make cpu metal server-metal      # binaries
make test test-sanitize          # hermetic units + ASan/UBSan
# model gates (direct; $M = GGUF, $V = pinned vectors):
build/kipp_test        --model         $M $V   # argmax + NMSE<=5e-5
build/kipp_test        --paged-cpu     $M $V   # scramble, bitwise
build/kipp_test        --multilogit    $M $V
build/kipp_test        --phase2-model  $M $V
build/kipp_test_metal  --metal-operators
build/kipp_test_metal  --phase3-metal  $M $V   # argmax + 1e-4
build/kipp_test_metal  --paged-metal   $M $V   # scramble, bitwise
build/kipp_test_metal  --multilogit-metal $M $V
make test-server                 # OpenAI API surface
tools/ops/verda_cuda_gate.sh     # CUDA gates, ephemeral cloud GPU
# mutation study (test-only build):
make build/kipp_test_fault
KIPP_FAULT=<1..4> build/kipp_test_fault --phase2-model $M $V
KIPP_FAULT=<1..4> build/kipp_test_fault --paged-cpu    $M $V
# benchmarks (each writes a self-describing bench/results/*.json):
python3 tools/bench.py ...             # decode/prefill by scheme and size
python3 bench/spec_bench.py --gate off # ungated speculation baseline
python3 bench/spec_bench.py --gate on  # adaptive-gated speculation
python3 bench/server_bench.py ...      # batch scaling
python3 bench/load_bench.py ...        # open-loop serving load
python3 bench/prefix_bench.py ...      # cross-request prefix reuse
python3 bench/ppl_bench.py ...         # WikiText-2 perplexity by scheme
# paper numbers (regenerate + verify the macro/data bindings):
make paper-data paper-check
\end{lstlisting}

\end{document}
